Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-12 paper/sections/metis-language-and-proofs.tex
[ paper ] fixing some typos and text added.

\begin{figure}
\begin{verbatim}
fof(a, axiom, p) .
fof(goal, conjecture, p) .
fof(subgoal_0, plain, p, inference (strip, [], [goal])) .
fof(negate_0_0, plain, ~ p, inference (negate, [], [subgoal_0])) .
fof(normalize_0_0, plain, ∼ p,
  inference (canonicalize, [], [negate_0_0])) .
fof(normalize_0_1, plain, p,
  inference (canonicalize, [], [a])) .
fof(normalize_0_2, plain, $false,
  inference (simplify, [], [normalize_0_0, normalize_0_1]))
cnf(refute_0_0, plain, $false,
  inference (canonicalize, [], [normalize_0_2])) .
\end{verbatim}
\caption{The \texttt{Metis}' \texttt{TSTP} derivation for the
problem $p\vdash p$.}
\label{fig:metis-proof-tstp}
\end{figure}

@@

For instance, for the propositional logic fragment without equality,
only three rules are valid from the \Metis' Logic Kernel:
\emph{assume}, \emph{axiom}, and \emph{resolve}. Nevertheless, as we
observed in \TSTP derivations, for such a propositional fragment,
\Metis uses in his deductions steps, six rules where only the
\emph{resolve} inference persists from the original list. In the
case of \resolve, it differs from its original version since, in
\TSTP, this rule has by definition an argument, a literal. The other
undocumented rules found in \TSTP derivations are: \canonicalize,
\clausify, \conjunct, \negate, \simplify, and \strip.

@@

This lack of the \Metis' documentation is an important matter for
us in the proof-reconstruction and certainty for other implementors
(see a comment related on the presenting of \name{Sledgehammer}
in \cite{paulson2007source} and \cite{Farber2015}). We faced
the ambiguity and lack of clearness of some rules like \canonicalize
or \simplify.
For example, we believe by applying these last
inferences, \Metis summarized many essential steps to deduce a
formula from a list a list of formulas as we see in the following
example.
\improvement{Add an example here to show a deduction where \Metis do
a lot between two steps.}
That inference' behavior is a drawback since, our approach
to reconstruct proofs needs take into account any assumption or
consideration as ordering for literals, application of
simplification theorems, reductions, among other valid
transformations.

@@

In this part, our main purpose consists of giving a response to
these rules looking at the \Metis' source code while inspecting a
representative set of solutions of \CPL problems
\cite{Prieto-Cubides2017} generated by \Metis.
This is an attempt to fulfill the \Metis's documentation, by
describing our each rule from our perspective. Later on, in section
\ref{sec:proof-reconstruction}, we present a formal description of
such rules.

@@

\textit{Clausification.} The \clausify rule transforms a
propositional formula in clausal normal form or \CNF, a conjunction
of clauses. Where a clause is the disjunction of zero or more
literals and a literal is an atom (positive literal) or a negation
of an atom (negative literal). This conversion is not unique, and
\Metis performs this transformation using its criteria.

@@

\[%\scalebox{0.9}{
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \bot$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \top$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \neg \varphi$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \varphi$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
%}
\]
\caption{Defintions of \canonicalize inference.}
\label{fig:conjunctive-disjunctive-simpl}
\end{figure}

@@

\textit{Simplification.}
The \simplify rule is inference that performs simplification of
definitions or tautologies over a list of derivations. This rule
transverses a list of formulas by applying different theorems (e.g.,
\emph{modus pones}, \emph{modus tollens}, or \emph{disjunctive
syllogism}) including the list of theorems used by \canonicalize in
Fig. \ref{fig:conjunctive-disjunctive-simpl} and the resolution
theorem of \resolve rule.

>>>>

\begin{figure}
\begin{verbatim}
fof(premise, axiom, p).
fof(goal, conjecture, p).
fof(subgoal_0, plain, p, inference(strip, [], [goal])).
fof(negate_0_0, plain, ~ p, inference(negate, [], [subgoal_0])).
fof(normalize_0_0, plain, ∼ p,
  inference(canonicalize, [], [negate_0_0])).
fof(normalize_0_1, plain, p,
  inference(canonicalize, [], [premise])).
fof(normalize_0_2, plain, $false,
  inference(simplify, [], [normalize_0_0, normalize_0_1]))
cnf(refute_0_0, plain, $false,
  inference(canonicalize, [], [normalize_0_2])).
\end{verbatim}
\caption{The \texttt{Metis}' \texttt{TSTP} derivation for the
problem $p\vdash p$.}
\label{fig:metis-proof-tstp}
\end{figure}

@@

For instance, for the propositional logic fragment without equality,
only three rules are valid from the \Metis' Logic Kernel:
\emph{assume}, \emph{axiom}, and \emph{resolve}. Nevertheless, as we
observed in \TSTP derivations, for such a propositional fragment,
\Metis uses in its deductions steps, six rules where only the
\emph{resolve} inference persists from the original list. In the
case of \resolve, it differs from its original version since, in
\TSTP, this rule has by definition an argument, a literal. The other
undocumented rules found in \TSTP derivations are: \canonicalize,
\clausify, \conjunct, \negate, \simplify, and \strip.

@@

This lack of the \Metis' documentation is an important matter for
us in the proof-reconstruction and certainty for other implementors
(see a comment related on the presenting of \name{Sledgehammer}
in \cite{paulson2007source} and \cite{Farber2015}). We faced
the ambiguity and lack of clearness of some rules like \canonicalize
or \simplify.
For example, we believe by applying these last
inferences, \Metis summarized many essential steps to deduce a
formula from a list of formulas.
\improvement{Add an example here to show a deduction where \Metis do
a lot between two steps.}
That inference's behavior is a drawback since, our approach
to reconstruct proofs needs take into account any assumption or
consideration as ordering for literals, application of
simplification theorems, reductions, among other valid
transformations.

@@

In this part, our main purpose consists of giving a response to
these rules looking at the \Metis' source code while inspecting a
representative set of solutions of \CPL problems
\cite{Prieto-Cubides2017} generated by \Metis.
This is an attempt to fulfill the \Metis's documentation, by
describing each rule from our perspective. Later on, in section
\ref{sec:proof-reconstruction}, we present a formal description of
such rules.

@@

\textit{Clausification.} The \clausify rule transforms a
propositional formula in clausal normal form or \CNF, a conjunction
of clauses. Where a clause is the disjunction of zero or more
literals and a literal is an atom (positive literal) or a negation
of an atom (negative literal). This conversion is not unique, and
\Metis performs this transformation using its criteria.
\improvement{We should mention that it was indeed a problem for the
proof-reconstruction? and refer the section next?}

@@

\[%\scalebox{0.9}{
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \bot$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \top$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \neg \varphi$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \varphi$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
%}
\]
\caption{Definitions for conjunction and disjunction simplification
in the \canonicalize inference.}
\label{fig:conjunctive-disjunctive-simpl}
\end{figure}

@@

\textit{Simplification.}
The \simplify rule is an inference that performs simplification of
definitions or tautologies over a list of derivations. This rule
transverses a list of formulas by applying different theorems (e.g.,
\emph{modus pones}, \emph{modus tollens}, or \emph{disjunctive
syllogism}) including the list of theorems used by \canonicalize in
Fig. \ref{fig:conjunctive-disjunctive-simpl} and the resolution
theorem of the \resolve rule.

### 2017-09-12 paper/sections/proof-reconstruction.tex
[ paper ] fixing some typos and text added.

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by Sultana in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is the proof
assistant, \Agda. We choose \Agda, but another proof assistant with
support of type theory and similar features like \Agda
(see a description of \Agda in section
\ref{sec:implementation}) is valid; our results do not depend on 
the proof assistant but the support of inductive data types, 
termination checking, and type-checking.

@@

In the rest of this section, we provide a formal description using
type theory to build definitions and theorems for the functions
necessary to emulate the \Metis' inference rules.

@@

\subsection{Recursion Theorems}

\improvement{Convention notation := for equivalence between propositions}
% \footnote{We use the \Agda standard library's implementation of the List data type.}.

\begin{definition}(Negative Normal Form (NNF))
Una forma normal negativa para la formula $\varphi$ es aquella
en la cual negaciones aparecen solamente en los literales y
la expresión se encuentra en términos unicamente
de los operadores booleanos ($\neg$, $\wedge$, $\vee$).
\end{definition}

Para transformar una formula $\varphi$ a su forma equivalente NNF,
aplicamos recursivamente las siguientes ecuaciones que definen
la función \name{nnf}.

@@

\begin{definition}
Una funcion recursiva se dice bien estructura, si el tamaño de los argumentos
con que con esta se define se reducen en tamaño por cada llamada a la
funcion.
\end{definition}

De la anterior definincion, sintacticamente la funcion nnf no es recursivamente
bien estructurado. El problem que no lo sea esta en nuestra traducion
de este tipo de funciones a la teoria de tipos.

Veamos otros ejemplos de funciones con el mismo problema que utilizaremos mas
adelante para formalizar la inferencia strip.

@@

Para traducir las funciones en un manera que esten bien definidas,
utilizamos el metodo propuesto por Ana Bove, el cual basicamente consiste

una funcion general recursiva $f:\sigma \to \tau$, su formalizacion en
teoria de tipos consiste en definir un predicado inductivo $Ind$,
$P:\sigma \to Ind$

En nuestro caso, el conjunto de los numeros servira como nuestro
conjunto inductivo y como segundo argumento para nueva funcion.
De manera, que la funcion g estara definida por recursion estructural en
su segundo termino.

Para aplicar la tecnica antes mencionada para nuestras funciones,
veamos su aplicación con la funcion \unshunt porque para las demas,
seguiremos un camino similar, f con tipo:

\[
\unshunt^{*} : PROP \to \mathcal{N} \to PROP
\]

@@

Then, we our final revision for the function $\unshunt$ is the
function$\unshunt^{+}$ with type $PROP \to PROP$
with a structured recursion,

@@

In the following section, our task will consist of a searching
appropriate functions like $rank$ to assure the termination of
our transformation functions like $nnf$.

@@

\emph{Splitting.}
We take a goal and generate subgoals that all in a conjunction
prove the goal.

@@

The function \mysplit performs such that strategy and it has the type
\[ \mysplit : \PROP \to \PROP
\]
\begin{equation}
\label{eq:split-definition}
\mysplit(φ) =
\begin{cases}
\unshunt(\mysplit(φ₁)) ∧ \unshunt(φ₁ ⇒ \mysplit(φ₂)
  & φ := φ₁ ∧ φ₂\\

@@

The theorem is the following.

@@

The \mysplit function returns a conjunction of the subgoal. From each
\TSTP derivation, we have information of the subgoals but no how they strictly appear in that conjunction. For that reason, with \Athena, we infer the exact order in the conjunction that coincides with the conjunction given by \mysplit.

\begin{theorem} For a list of subgoals $s_0, \cdots, s_1$ with their respective theorem $Γ ⊢ s_i$, such that $\bigwedge_i s_i ⇒ \texttt{goal}$, then $Γ ⊢ \texttt{goal}$ 
\end{theorem}

>>>>

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by Sultana in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is the proof
assistant, \Agda. We choose \Agda, but another proof assistant with
support of Type Theory and similar features like \Agda
(see a description of \Agda in section
\ref{sec:implementation}) is valid; our results do not depend on
the proof assistant but the support of inductive data types,
termination checking, and type-checking.

@@

In the rest of this section, we provide a formal description using
Type Theory to build definitions and theorems for the functions
necessary to emulate the \Metis' inference rules.

@@

\subsection{Rules as Theorems}

Main algorithms in \CPL need the propositions in some of their normal
forms. For instance, from the negative normal form
(henceforth, \abbre{NNF}), we can derive the conjunctive normal form,
and the disjunctive normal form as well.
Therefore, we start giving the definition of \abbre{NNF}, one
function definition, \name{nnf} and a refinement of it.
We will use it later for instance in the description of \canonicalize
rule. The following treatment to this function, \name{nnf},
applies for other conversion functions and theorems defined in
\cite{AgdaProp,AgdaMetis}.
We use the notation convention $:=$ for equivalence between
propositions. The \PROP inductive data type is defined as usual
based on the definition from section \ref{sec:preliminaries} and
implemented in \Agda with \name{PropFormula} in section
\ref{sec:implementation}.

\subsubsection{\CPL Functions}
\hspace*{5cm}\\[3mm]
\emph{NNF.}
A negative normal form for the formula φ is one in which negations
appear only in the literals and the expression is in terms only of
the boolean operators (¬, ∧, ∨).

To transform a formula φ to its equivalent \abbre{NNF} form, we
apply recursively the following equations. The function type for
\name{nnf} is $\PROP \to \PROP$.

@@

A recursive function is a well-structured recursion, if the size of
the arguments with which it is defined decrease in size for each
call to the function in the definition. That guarantee termination
of the function, and its translation to constructive Type Theory.

But we need to improve the definition for \name{nnf} in order to have
a version of a well-structured recursion.
We show a strategy to achieve that with a smaller example function,
\unshunt, that we will use forward to formalize the inference
\name{strip}.

@@

To translate the functions in a way that is well defined,
we used the method proposed by Bove and Capettra in
\cite{Bove2005}, which basically consists of

We take a general recursive function $f: \sigma \to \tau$.
Its formalization in Type Theory will define an inductive
predicate $Ind$, $ P: \sigma \to Ind$.

In our case, the set of natural numbers will serve as our
inductive set $Ind$ and as second argument for new function.
Thus, the function $\unshunt$ will be defined by structural
recursion in his second term.

The new proposed $\unshunt^{*}$ has type,

\[
\unshunt^{*} : PROP \to \mathbb{N} \to PROP.
\]

@@

Then, we our final review for the function $\unshunt$ is the
function$\unshunt^{+}$ with type $PROP \to PROP$
with a structured recursion,

@@

In the following section, our task will consist of a searching
appropriate functions like $rank$ to assure the termination for transformation functions like $nnf$.

@@

\emph{Splitting.}
To prove a goal, we can prove smaller subgoals and combine their
proofs to achieve the final target, a proof of the initial goal.

@@

The function \mysplit has the type
\[ \mysplit : \PROP \to \PROP
\]
\begin{equation}
\label{eq:split-definition}
\mysplit(φ) =
\begin{cases}
\unshunt(\mysplit(φ₁)) ∧ \unshunt(φ₁ ⇒ \mysplit(φ₂)
  & φ := φ₁ ∧ φ₂\\

@@

@@

The \mysplit function returns a conjunction of the subgoal. From each
\TSTP derivation, we have information of the subgoals but no how they strictly appear in that conjunction. For that reason, with \Athena, we infer the exact order in the conjunction that coincides with the output by \mysplit.

\begin{theorem} For a list of subgoals $s_0, \cdots, s_1$ with their respective theorem $Γ ⊢ s_i$, such that $\bigwedge_i s_i ⇒ \texttt{goal}$, then $Γ ⊢ \texttt{goal}$.
\end{theorem}
